---
title: "Compare future ENA and S_t among IMAGE scenarios"
author: "Andy Purvis"
date: "24/02/2021"
output:
  html_document: 
    df_print: paged
    code_folding: hide
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
  pdf_document:
    df_print: kable
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Comparing effective natural area and extinctions projected for the future among the IMAGE scenarios

The results of the relaxation analyses for the IMAGE scenarios include some big surprises. I am therefore mapping the differences among scenarios in terms of the effective natural area projected for the year 2050 and the numbers of extinctions projected for then.

```{r, load, echo=FALSE, cache=FALSE}
library(checkpoint)
checkpoint("2021-02-16")

start.time <- Sys.time()
library(raster)
library(scales)
library(abind)
library(gplots)
library(colorspace)
library(maps)
library(ncdf4) # package for netcdf manipulation, suggested by https://rpubs.com/boyerag/297592
library(rgdal) # package for geospatial analysis, suggested by https://rpubs.com/boyerag/297592
library(ggplot2) # package for plotting, suggested by https://rpubs.com/boyerag/297592
source("00_functions.R")
source("convert_iam_data.R")
```
## Reading in ENA raster stacks from IMAGE scenarios

In 2030, the GDN scenario sees markedly more effective natural area in many African cells than in HPR, but less elsewhere.

In 2030, BID has markedly higher effective natural area than CRJ in some endemic-rich regions (notably the Andean chain and East African rift system); but HPR also has higher effective natural area than CRJ in these areas. 

```{r, read_alsu_stacks, echo=FALSE, cache=TRUE, out.width="100%"}
instem <- "output/ena"
A_BID <- stack(file.path(instem, "ENA-IMAGE-RCPref_SSP2_BID-06Feb2020_FinalMask.nc"))
A_CRJ <- stack(file.path(instem, "ENA-IMAGE-RCPref_SSP2_CRJ-06Feb2020_FinalMask.nc"))
A_GDN <- stack(file.path(instem, "ENA-IMAGE-RCPref_SSP2_GDN-06Feb2020_FinalMask.nc"))
A_HPR <- stack(file.path(instem, "ENA-IMAGE-RCPref_SSP2_HPR-06Feb2020_FinalMask.nc"))
A_NRS <- stack(file.path(instem, "ENA-IMAGE-RCPref_SSP2_NRS-06Feb2020_FinalMask.nc"))
A_THC <- stack(file.path(instem, "ENA-IMAGE-RCPref_SSP2_THC-06Feb2020_FinalMask.nc"))
A_WAC <- stack(file.path(instem, "ENA-IMAGE-RCPref_SSP2_WAC-06Feb2020_FinalMask.nc"))

y <- 5 # 2030 = 5, 2050 = 9

par(mar = c(5, 4, 4, 2) + 0.1)
plot(subset(A_GDN, y) - subset(A_HPR, y), main = "GDN - HPR: Effective natural area in 2030")
plot(subset(A_BID, y) - subset(A_CRJ, y), main = "BID - CRJ: Effective natural area in 2030")
plot(subset(A_CRJ, y) - subset(A_HPR, y), main = "CRJ - HPR: Effective natural area in 2030")

```

## Reading in extinction arrays from projections

Compared with HPR, GDN sees more endemic species survive in 2050 in parts of Africa, Madagascar and some parts of the S E Asian archipelagoes; but HPR sees more endemic species survive than GDN in parts of the New World, Australia and the E Asian mainland.

BID retains more endemics in the Andean chain in 2050 than does CRJ; but so does HPR.

```{r, read_S_t_arrays, echo=FALSE, cache=TRUE, fig.show="hold", out.width="50%"}
instem <- "output/arrays"
S_BID <- readRDS(file.path(instem, "asluBID-06Feb2020_FinalMask.nc.Rds"))
S_CRJ <- readRDS(file.path(instem, "asluCRJ-06Feb2020_FinalMask.nc.Rds"))
S_GDN <- readRDS(file.path(instem, "asluGDN-06Feb2020_FinalMask.nc.Rds"))
S_HPR <- readRDS(file.path(instem, "asluHPR-06Feb2020_FinalMask.nc.Rds"))
S_NRS <- readRDS(file.path(instem, "asluNRS-06Feb2020_FinalMask.nc.Rds"))
S_THC <- readRDS(file.path(instem, "asluTHC-06Feb2020_FinalMask.nc.Rds"))
S_WAC <- readRDS(file.path(instem, "asluWAC-06Feb2020_FinalMask.nc.Rds"))

y <- 21 # 41 = 2050, 21 = 2030

par(mar = c(4, 4, .1, .1))
diff <- S_GDN[,,y] - S_HPR[,,y]
image(t(diff * (diff>0)), ylim=c(1,0), bty="n", xaxt="n", yaxt="n", main="GDN has more endemics in 2030 than HPR")
image(t(-diff * (diff < 0)), ylim=c(1,0), bty="n", xaxt="n", yaxt="n", main="HPR has more endemics in 2030 than GDN")

diff <- S_BID[,,y] - S_CRJ[,,y]
image(t(diff * (diff>0)), ylim=c(1,0), bty="n", xaxt="n", yaxt="n", main="BID has more endemics in 2030 than CRJ")
image(t(-diff * (diff < 0)), ylim=c(1,0), bty="n", xaxt="n", yaxt="n", main="CRJ has more endemics in 2030 than BID")

diff <- S_CRJ[,,y] - S_HPR[,,y]
image(t(diff * (diff>0)), ylim=c(1,0), bty="n", xaxt="n", yaxt="n", main="CRJ has more endemics in 2030 than HPR")
image(t(-diff * (diff < 0)), ylim=c(1,0), bty="n", xaxt="n", yaxt="n", main="HPR has more endemics in 2030 than CRJ")

```
