---
title: "Comparing tiff and netCDF IAM data"
author: "Andy Purvis"
date: "10/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Comparing tiff and netCDF IAM data

```{r, load, echo=FALSE, cache=TRUE}
start.time <- Sys.time()
library(raster)
library(scales)
library(abind)
library(gplots)
library(colorspace)
library(maps)
library(ncdf4) # package for netcdf manipulation, suggested by https://rpubs.com/boyerag/297592
library(rgdal) # package for geospatial analysis, suggested by https://rpubs.com/boyerag/297592
library(ggplot2) # package for plotting, suggested by https://rpubs.com/boyerag/297592
source("00_functions.R")
```

## Load LUH2 effective natural area
This map shows the effective area of natural habitat per 2$^\circ$ grid cell for the year 2010, as estimated previously from just the LUH2 data used in De Palma et al. (in prep.).
```{r read-luh2}
luh2_ena <- stack("output/ENA_LUH2.nc")
plot(luh2_ena, 233, main = "LUH2 ENA in 2010")
```

## Load tiff for same IAM/scenario combination as used in previous examples
Unfortunately, there has been a "Chinese whispers" situation with the organisation of the tifs. Rather than having a stack for each land use, I have been provided with 12 snapshots for each decade. That will cause a problem if I have to construct the time series myself, but does not cause any particular problems for estimating the effective natural area for the year 2010.

Each of the rasters is mapped below. Note that there are only 11 rasters: none was provided for abn_crop2Gbioen. It is unclear whether that's an omission or has happened because this year, scenario and/or IAM has none of it. None is expected prior to 2020 anyway, so its lack should not affect the comparison with LUH2 and the netCDF data.

```{r read-iam, echo=FALSE}
stem <- "data/IAM_FinalScenarios_rasters"
iamID <- "MAGPIE_rasters"
scenario <- "GDN"
year <- "2010"
cell_areas <- raster(file.path(stem,iamID,scenario,"MAGPIE_GDN_all_area.tif"))

# Read in the twelve rasters
# Multiply cell land-use fractions by cell area before summing into 2-degree grids
crop_other <- aggregate(raster(file.path(stem,iamID,scenario,year,"MAGPIE_GDN_share_cropland_other_2010.tif")) * cell_areas, 
                 fact = 4, fun = sum)
plot(crop_other, main = "crop_other")

crop_2Gbioen <- aggregate(raster(file.path(stem,iamID,scenario,year,"MAGPIE_GDN_share_cropland_2Gbioen_2010.tif")) * cell_areas, 
                 fact = 4, fun = sum)
plot(crop_2Gbioen, main = "crop_2Gbioen")

grassland <- aggregate(raster(file.path(stem,iamID,scenario,year,"MAGPIE_GDN_share_grassland_2010.tif")) * cell_areas, 
                 fact = 4, fun = sum)
plot(grassland, main = "grassland")

forest_unmanaged <- aggregate(raster(file.path(stem,iamID,scenario,year,"MAGPIE_GDN_share_forest_unmanaged_2010.tif")) * cell_areas, 
                 fact = 4, fun = sum)
plot(forest_unmanaged, main="forest_unmanaged")

forest_managed <- aggregate(raster(file.path(stem,iamID,scenario,year,"MAGPIE_GDN_share_forest_managed_2010.tif")) * cell_areas, 
                 fact = 4, fun = sum)
plot(forest_managed, main="forest_managed")

restored <- aggregate(raster(file.path(stem,iamID,scenario,year,"MAGPIE_GDN_share_restored_2010.tif")) * cell_areas, 
                 fact = 4, fun = sum)
plot(restored, main="restored")


built_up <- aggregate(raster(file.path(stem,iamID,scenario,year,"MAGPIE_GDN_share_built-up_2010.tif")) * cell_areas, 
                 fact = 4, fun = sum)
plot(built_up, main="built_up")


other <- aggregate(raster(file.path(stem,iamID,scenario,year,"MAGPIE_GDN_share_other_2010.tif")) * cell_areas, 
                 fact = 4, fun = sum)
plot(other, main="other")


abn_cropland_other <- aggregate(raster(file.path(stem,iamID,scenario,year,"MAGPIE_GDN_share_abn_cropland_other_2010.tif")) * cell_areas, 
                 fact = 4, fun = sum)
plot(abn_cropland_other, main="abn_cropland_other")


# There is no raster for abn_cropland_2Gbioen!


abn_grassland <- aggregate(raster(file.path(stem,iamID,scenario,year,"MAGPIE_GDN_share_abn_grassland_2010.tif")) * cell_areas, 
                 fact = 4, fun = sum)
plot(abn_grassland, main="abn_grassland")


abn_forest_managed <- aggregate(raster(file.path(stem,iamID,scenario,year,"MAGPIE_GDN_share_abn_forest_managed_2010.tif")) * cell_areas, 
                 fact = 4, fun = sum)
plot(abn_forest_managed, main="abn_forest_managed")

```

## Calculating effective natural area by combining these rasters with their corresponding rescaled BII estimates.
```{r, calculate-effective-natural-area, echo=FALSE, cache=TRUE}
# Read in file of BII values
bii_coefficients <- read.csv("output/Rescaled_BII_values_for_IAM_classes.csv")
print("Read the bii file in")

# Choose which of two rescalings to use
which_bii <- "BII_humdom0"
if(which_bii %in% c("BII_crop0", "BII_humdom0") == FALSE){
  stop("Invalid value for which_bii")
}

if (which_bii == "BII_humdom0"){
  bii <- bii_coefficients$BII_humdom0
}else{
  bii <- bii_coefficients$BII_crop0
}

# Multiply each raster by its BII and add the products together
effective_natural_area <- 
  crop_other * bii[1] +
  crop_2Gbioen * bii[2] +
  grassland * bii[3] +
  forest_unmanaged * bii[4] +
  forest_managed * bii[5] +
  restored * bii[6] +
  other * bii[7] +
  built_up * bii[8] +
  abn_cropland_other * bii[9] +
  abn_grassland * bii[11] +
  abn_forest_managed * bii[12]
plot(effective_natural_area, main="ENA derived from tif")
map(add=TRUE)

if (require(ncdf4)) {	
     rnc <- writeRaster(effective_natural_area, filename=file.path("output", paste("ENA_from_tif_", which_bii, scenario, sep="_")), format="CDF", overwrite=TRUE) 
}
tif_ENA <- effective_natural_area
```
## Compare tif-derived and netCDF-derived estimates of effective natural area in 2010
The map of netCDF-derived estimates of effective natural area is identical, with exactly the same cells having missing values.
```{r, compare-effective-natural-area, echo=FALSE, cache=TRUE}
netCDF_ENA <- subset(stack(file.path("output", "ENA_BII_humdom0_BendingTheCurve30by30-LCproj-MAgPIE-RCPref_SSP2_GDN-06Feb2020_FinalMask.nc")), 1)
plot(netCDF_ENA, main="ENA derived from netCDF")
map(add=TRUE)

plot(getValues(tif_ENA)~getValues(netCDF_ENA), ylab="From TIF", xlab="From netCDF")

#The following condition will be true if and only if the same cells have missing values
sum(is.na(getValues(tif_ENA))) == sum(is.na(getValues(tif_ENA)) & is.na(getValues(netCDF_ENA)))
```
