---
title: "Comparing tiff and netCDF IAM data"
author: "Andy Purvis"
date: "10/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Comparing tiff and netCDF IAM data

```{r, load, echo=FALSE, cache=TRUE}
start.time <- Sys.time()
library(raster)
library(scales)
library(abind)
library(gplots)
library(colorspace)
library(ncdf4) # package for netcdf manipulation, suggested by https://rpubs.com/boyerag/297592
library(rgdal) # package for geospatial analysis, suggested by https://rpubs.com/boyerag/297592
library(ggplot2) # package for plotting, suggested by https://rpubs.com/boyerag/297592
source("00_functions.R")
```

## Load LUH2 effective natural area
This map shows the effective area of natural habitat per 2$^\circ$ grid cell for the year 2010, as estimated previously from just the LUH2 data used in De Palma et al. (in prep.).
```{r read-luh2}
luh2_ena <- stack("output/ENA_LUH2.nc")
plot(luh2_ena, 233, main = "LUH2 ENA in 2010")
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
