---
title: "Compare IAM data with LUH2 and rescale LUH2"
author: "Andy Purvis"
date: "09/02/2021"
output:
  html_document: 
    df_print: paged
    code_folding: hide
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
  pdf_document:
    df_print: kable
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Comparison of IAM data with LUH2 data and rescaling of LUH2 data
Although IAM data and LUH2 data are not expected to be entirely congruent during the years for which both are available, they are expected to be strongly correlated -- both are conditioned on the same current global pattern of land use, albeit slightly different interpretations of that pattern. The logic of the dynamic species-area model requires that the history of land use (provided by LUH2) melds reasonably well with the scenario-dependent future of land use (provided by the IAMs). The approach taken here is to rescale LUH2 data so that it exactly matches the IAM data for the year 2010. But before that rescaling, it is sensible to test that the correlation across grid cells between IAM and LUH2 data is reasonable.

```{r, load, echo=FALSE, cache=TRUE}
start.time <- Sys.time()
library(raster)
library(scales)
library(abind)
library(gplots)
library(colorspace)
library(maps)
library(ncdf4) # package for netcdf manipulation, suggested by https://rpubs.com/boyerag/297592
library(rgdal) # package for geospatial analysis, suggested by https://rpubs.com/boyerag/297592
library(ggplot2) # package for plotting, suggested by https://rpubs.com/boyerag/297592
source("00_functions.R")
```

```{r, set-params, echo=FALSE, cache=TRUE}
bii_sec <- mean(c(0.9335480, 1, 0.7577225)) # mean = 0.8970902, following De Palma et al. (in prep.)
luh2_years <- seq(850, 2015, by = 5)
iam_years <- seq(2010, 2100, by = 5)


```
## Loading in LUH2 data, calculating effective natural area and plot map for 2010

```{r, load-luh2, echo=FALSE, cache=TRUE}
# Read in historical LUH2 data
cell_areas <- aggregate(raster("output/Waldron_cell_areas.nc"), fact = 4, fun=sum) # from one of the IAM netCDFs, saved earlier
pri <- stack("data/historical-primary.tif") # producted for De Palma et al. (in prep.)
sec <- stack("data/historical-secondary.tif") # produced for De Palma et al. (in prep.)

# Calculate effective natural area
luh2_fraction <- clamp(pri + bii_sec * sec, lower=0, upper=1, useValues=TRUE)
luh2_area <- luh2_fraction * cell_areas

plot(luh2_area, y=233, main="LUH2 effective natural area in 2010") # 2010
map(add=TRUE)
```

## Compare with IAM data for 2010
• N.B. New plots added since this text was updated!
The first plot shows the effective natural area in 2010 according to the IAM used in the earlier examples. The second plot shows the equivalent map for the LUH2 data, and the third shows the difference (as LUH2 - IAM). Note that many cells have large differences. Note also that the top two maps have different sets of pixels (most obvious when looking at Greenland). The fourth plot shows the cellwise correlation between the two data sets with the 1:1 line. If NAs are treated as zeroes, the correlation is 0.887; if left as NAs it is 0.72 -- both markedly lower than in the Dasgupta work (where the correlation was 0.976). This is disconcertingly low, and requires checking before going on to harmonise the effective area data between the two sources. 

• Checking the BII coefficients for the IAM land uses revealed no errors.
• Although it cannot be the reason for the low correlation, the the different cell sets is interesting, and it will be necessary to check whether the masks are the same in all scenarios, IAMs and years.

The last three maps show the masks: first, cells that are NA in LUH2 only; then cells that are NA in the IAM data only; and lastly cells that are NA in both data sets.

```{r, compare-iam, echo=FALSE, cache=TRUE}
iam <- stack(file.path("output", "ENA_BII_humdom0_BendingTheCurve30by30-LCproj-MAgPIE-RCPref_SSP2_GDN-06Feb2020_FinalMask.nc"))
iam_crop0 <- stack(file.path("output", "ENA_BII_crop0_BendingTheCurve30by30-LCproj-MAgPIE-RCPref_SSP2_GDN-06Feb2020_FinalMask.nc"))

iam_2010 <- subset(iam, 1)
iam_crop0_2010 <- subset(iam_crop0, 1)
luh2_2010 <- subset(luh2_area, 233)
par(mar = c(5,4,4,2)+0.1) # Default

plot(iam_2010, main="IAM humdom0 effective natural area, 2010")
map(add=TRUE)
plot(iam_crop0_2010, main ="IAM crop0 effective natural area, 2010")
map(add=TRUE)
plot(luh2_2010, main = "LUH2 effective natural area, 2010")
map(add=TRUE)
plot(luh2_2010 - iam_2010, main = "LUH2 - IAM humdom0 in 2010")
map(add=TRUE)
plot(luh2_2010 - iam_crop0_2010, main = "LUH2 - IAM crop0 in 2010")
map(add=TRUE)

bigdiff_humdom0 <- ((iam_2010 - luh2_2010)^2 > 1)
plot(bigdiff_humdom0, main = "Absolute difference > 1 Mha: BII_humdom0")
bigdiff_crop0 <- ((iam_crop0_2010 - luh2_2010)^2 > 1)
plot(bigdiff_crop0, main = "Absolute difference > 1 Mha: BII_crop0")

y1 <- getValues(luh2_2010)
y2 <- getValues(iam_2010)
valid <- !is.na(y1 * y2)
print(paste("Cell-wise correlation (humdom0) excluding NAs:", cor(y1[valid], y2[valid])))

y1[is.na(y1)] <- 0
y2[is.na(y2)] <- 0
print(paste("Cell-wise correlation (humdom0) with NAs set to zero:", cor(y1, y2)))


plot(y1~y2, ylab="LUH2 in 2010", xlab="IAM humdom0 in 2010")
abline(a=0,b=1, col="red")

y1 <- getValues(luh2_2010)
y2 <- getValues(iam_crop0_2010)
valid <- !is.na(y1 * y2)
print(paste("Cell-wise correlation (crop0) excluding NAs:", cor(y1[valid], y2[valid])))

y1[is.na(y1)] <- 0
y2[is.na(y2)] <- 0
print(paste("Cell-wise correlation (crop0) with NAs set to zero:", cor(y1, y2)))


plot(y1~y2, ylab="LUH2 in 2010", xlab="IAM crop0 in 2010")
abline(a=0,b=1, col="red")


# 709 cells are NA in the IAMs but have values in LUH2 - where are they?
na_in_iam_only <- (is.na(iam_2010) & !is.na(luh2_2010))
na_in_luh2_only <- (is.na(luh2_2010) & !is.na(iam_2010))
na_in_both <- (is.na(luh2_2010) & is.na(iam_2010))
plot(na_in_iam_only, main = "NA in IAM only")
map(add=TRUE)
plot(na_in_luh2_only, main = "NA in LUH2 only")
map(add=TRUE)
plot(na_in_both, main = "NA in both IAM and LUH2")
map(add=TRUE)

```

```{r, write_luh2_ENA, echo=FALSE, cache=TRUE}
if (require(ncdf4)) {	
     rnc <- writeRaster(luh2_area, filename=file.path("output", "ENA_LUH2.nc"), format="CDF", overwrite=TRUE) 
}

```