---
title: "Compare future ENA and S_t among IMAGE scenarios"
author: "Andy Purvis"
date: "24/02/2021"
output:
  pdf_document:
    df_print: kable
    toc: true
    toc_depth: 2
    number_sections: true
  html_document: 
    df_print: paged
    code_folding: hide
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Comparing effective natural area and extinctions projected for the future among the IMAGE scenarios

The results of the relaxation analyses for the IMAGE scenarios include some big surprises. I am therefore mapping the differences among scenarios in terms of the effective natural area projected for the year 2050 and the numbers of extinctions projected for then.

```{r, load, echo=FALSE, cache=FALSE}
library(checkpoint)
checkpoint("2021-02-16")

start.time <- Sys.time()
library(raster)
library(scales)
library(abind)
library(gplots)
library(colorspace)
library(maps)
library(ncdf4) # package for netcdf manipulation, suggested by https://rpubs.com/boyerag/297592
library(rgdal) # package for geospatial analysis, suggested by https://rpubs.com/boyerag/297592
library(ggplot2) # package for plotting, suggested by https://rpubs.com/boyerag/297592
source("00_functions.R")
source("convert_iam_data.R")
```
## Reading in ENA raster stacks from IMAGE scenarios

In 2050, the GDN scenario sees markedly more effective natural area in many African cells than in HPR, but less elsewhere.

```{r, read_alsu_stacks, echo=FALSE, cache=TRUE, out.width="100%"}
instem <- "output/ena"
A_BID <- stack(file.path(instem, "ENA-IMAGE-RCPref_SSP2_BID-06Feb2020_FinalMask.nc"))
A_CRJ <- stack(file.path(instem, "ENA-IMAGE-RCPref_SSP2_CRJ-06Feb2020_FinalMask.nc"))
A_GDN <- stack(file.path(instem, "ENA-IMAGE-RCPref_SSP2_GDN-06Feb2020_FinalMask.nc"))
A_HPR <- stack(file.path(instem, "ENA-IMAGE-RCPref_SSP2_HPR-06Feb2020_FinalMask.nc"))
A_NRS <- stack(file.path(instem, "ENA-IMAGE-RCPref_SSP2_NRS-06Feb2020_FinalMask.nc"))
A_THC <- stack(file.path(instem, "ENA-IMAGE-RCPref_SSP2_THC-06Feb2020_FinalMask.nc"))
A_WAC <- stack(file.path(instem, "ENA-IMAGE-RCPref_SSP2_WAC-06Feb2020_FinalMask.nc"))

y2050 <- 9

par(mar = c(5, 4, 4, 2) + 0.1)
plot(subset(A_GDN, y2050) - subset(A_HPR, y2050), main = "GDN - HPR: Effective natural area in 2050")

```

## Reading in extinction arrays from projections

Compared with HPR, GDN sees more endemic species survive in 2050 in parts of Africa, Madagascar and some parts of the S E Asian archipelagoes; but HPR sees more endemic species survive than GDN in parts of the New World, Australia and the E Asian mainland.

```{r, read_S_t_arrays, echo=FALSE, cache=TRUE, fig.show="hold", out.width="50%"}
instem <- "output/arrays"
S_BID <- readRDS(file.path(instem, "asluBID-06Feb2020_FinalMask.nc.Rds"))
S_CRJ <- readRDS(file.path(instem, "asluCRJ-06Feb2020_FinalMask.nc.Rds"))
S_GDN <- readRDS(file.path(instem, "asluGDN-06Feb2020_FinalMask.nc.Rds"))
S_HPR <- readRDS(file.path(instem, "asluHPR-06Feb2020_FinalMask.nc.Rds"))
S_NRS <- readRDS(file.path(instem, "asluNRS-06Feb2020_FinalMask.nc.Rds"))
S_THC <- readRDS(file.path(instem, "asluTHC-06Feb2020_FinalMask.nc.Rds"))
S_WAC <- readRDS(file.path(instem, "asluWAC-06Feb2020_FinalMask.nc.Rds"))

y2050 <- 41

par(mar = c(4, 4, .1, .1))
diff <- S_GDN[,,y2050] - S_HPR[,,y2050]
image(t(diff * (diff>0)), ylim=c(1,0), bty="n", xaxt="n", yaxt="n", main="GDN has more endemics in 2050 than HPR")
image(t(-diff * (diff < 0)), ylim=c(1,0), bty="n", xaxt="n", yaxt="n", main="HPR has more endemics in 2050 than GDN")

```
