---
title: "Converting all IAM land-use data to ENA"
author: "Andy Purvis"
date: "11/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Converting all IAM land-use data to Effective Natural Area

Land-use data from four IAMs have been provided. I have rationalised the folder structure manually to facilitate looping. Specifically, the GLOBIOM files were in a folder called "GLOBIOM" within another folder called "GLOBIOM", so I removed the otherwise-empty folder; and the AIM folder contained multiple versions, with one folder called "AIM final version use this" -- I renamed this latter folder "AIMfinal" and discarded all the other AIM files and folders. Three of the folders contain eight files, whereas IMAGE has just seven (it has no NOBIOD file).

```{r, load, echo=FALSE, cache=FALSE}
library(raster)
library(scales)
library(abind)
library(gplots)
library(colorspace)
library(maps)
library(ncdf4) # package for netcdf manipulation, suggested by https://rpubs.com/boyerag/297592
library(rgdal) # package for geospatial analysis, suggested by https://rpubs.com/boyerag/297592
library(ggplot2) # package for plotting, suggested by https://rpubs.com/boyerag/297592
source("convert_iam_data.R")

```

The code loops through the four IAMs, processing each file within their respective folders (seven or eight per IAM). For each, it writes a new raster stack and (optionally) a pdf report. For the first data set within each folder, it also maps the signed difference between IAM and LUH2 for the year 2010 and plots the cell-wise correlation. (There is no need to do this for all the files in the folder, as the others have identical data for 2010, differing only after 2020.)

AIM and IMAGE provide effective natural area estimates that agree much more closely with LUH2 than do GLOBIOM and MAgPIE.

```{r, set-up, echo=FALSE, cache=TRUE, fig.show="hold", out.width="50%"}
par(mar = c(4, 4, .1, .1))
luh2_2010 <- raster("output/ENA_LUH2_2010.nc")
vl <- getValues(luh2_2010)

which_bii <- "BII_humdom0"
iam_set <- c("AIMfinal", "GLOBIOM", "IMAGE", "MAGPIE")
scenarios <- c("BID", "CRJ", "GDN", "HPR", "NOBIOD", "NRS", "THC", "WAC")
in_stem <- "data/IAM outputs for 30 by 30 scenarios"
ena_stem <- "output/ena"
pdf_stem <- "reports/Converting IAMs to ENA"

for (iam in iam_set){
  to_convert <- sort(list.files(file.path(in_stem, iam)))
  for (i in 1:length(to_convert)){
    print(paste("Converting", to_convert[i], "..."))
    iam_path <- file.path(in_stem, iam, to_convert[i])
    ena_path <-file.path(ena_stem, gsub("BendingTheCurve30by30-LCproj-", "ENA-", to_convert[i]))
    pdf_name <- gsub("BendingTheCurve30by30-LCproj-", "Converting-", to_convert[i])
    pdf_name <- gsub(".nc", ".pdf", pdf_name)
    pdf_path <- file.path(pdf_stem, pdf_name)
    
    iam_2010 <- convert_iam_data(iam_path, which_bii, ena_path, pdf_path)
    
    if (i == 1){
      # Map and compare
      plot(luh2_2010 - iam_2010, main = paste("LUH2 -", iam))
      vi <- getValues(iam_2010)
      keep <- !is.na(vl*vi)
      plot(luh2_2010, iam_2010, ylab=iam, xlab="LUH2") 
      text(0, 4.5, paste("r =", round(cor(vi[keep], vl[keep]), 3)), adj=0)
      abline(a=0,b=1, col="red")
    }
  }
  
}

```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
