---
title: "Digging into surprising IMAGE results"
author: "Andy Purvis"
date: "25/02/2021"
output:
  html_document: 
    df_print: paged
    code_folding: hide
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
  pdf_document:
    df_print: kable
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Digging into surprising IMAGE results

The IMAGE-based scenarios have surprising and non-intuitive results. Specifically, the HPR scenario (for Harsh Political Reality) reduces extinctions of endemics by more than any other scenario, despite being conceived of as the least biodiversity-oriented of the action scenarios (NRS is a non-action counterfactual, in which protected area coverage is not extended).

Given how surprising this result is, it is sensible to check nothing has gone wrong in the NHM processing of the input files; e.g., mixing up two input files could potentially drive the patterns we see. This markdown examines carefully each step in the processing, focusing on three scenarios that Anthony Waldron explained form a sort of linear progression in terms of how biodiversity-friendly they are intended to be: HPR, CRJ and BID.

Steps to investigate:
  
  • Do the ENA files for the scenarios agree at 2020, as they should; and, if not, at what step in the processing did they become different?
  
  • Can I trace from the input netCDFs to the ENA files, or did something get, e.g., misnamed?
  
## About the scenarios

Anthony Waldron characterised the three scenarios as follows:

• BID = optimum 30% to minimise extinction under constraint of existing, inefficient PA system

• HPR = optimum 30% under constraint that any land that will be used for agriculture by 2050 cannot be a PA

• CRJ = a compromise between those two (optimum 20% for biodiversity, then avoid agricultural land for the remaining 10% to make up 30%...I believe)

Hence, by design, they are expected to lie on a continuum in terms of their mitigation of extinctions.

```{r, load, echo=FALSE, cache=FALSE}
library(checkpoint)
checkpoint("2021-02-16")

start.time <- Sys.time()
library(raster)
library(scales)
library(abind)
library(gplots)
library(colorspace)
library(maps)
library(ncdf4) # package for netcdf manipulation, suggested by https://rpubs.com/boyerag/297592
library(rgdal) # package for geospatial analysis, suggested by https://rpubs.com/boyerag/297592
library(ggplot2) # package for plotting, suggested by https://rpubs.com/boyerag/297592
source("00_functions.R")
source("convert_iam_data.R")
```

## Comparing ENA files for the three scenarios every five years from 2010 to 2030

The effective natural area stacks calculated previously are the same for BID, CRJ and HPR (beyond rounding error) in 2010, 2015 and 2020 -- as they should be -- but have begun to diverge in the 2025 data. 

```{r compare-ENA, echo=FALSE, cache=TRUE}
stem <- "output/ena"
ENA_BID <- stack(file.path(stem, "ENA-IMAGE-RCPref_SSP2_BID-06Feb2020_FinalMask.nc"))
ENA_CRJ <- stack(file.path(stem, "ENA-IMAGE-RCPref_SSP2_CRJ-06Feb2020_FinalMask.nc"))
ENA_HPR <- stack(file.path(stem, "ENA-IMAGE-RCPref_SSP2_HPR-06Feb2020_FinalMask.nc"))

year <- seq(2010,2100,by=5)

par(mfrow=c(1,2))

for (i in 1:5){
  #Loop through first 5 layers of each stack
  sd <- sd(getValues(subset(ENA_BID, i) - subset(ENA_CRJ, i)), na.rm=TRUE)
  plot(subset(ENA_BID, i) - subset(ENA_CRJ, i), main = paste("BID - CRJ: ENA in", year[i]), sub = paste("Standard deviation =", round(sd, 4)))
  sd <- sd(getValues(subset(ENA_CRJ, i) - subset(ENA_HPR, i)), na.rm=TRUE)
  plot(subset(ENA_CRJ, i) - subset(ENA_HPR, i), main = paste("CRJ - HPR: ENA in", year[i]), sub = paste("Standard deviation =", round(sd, 4)))
}



```

## Checking the netCDF files shared with me

The netCDF files shared with me are in data/IAM outputs for 30 by 30 scenarios/IMAGE. There are 12 land-use classes, of which 3 are relevant:

• Layer 4: forest_unmanaged
• Layer 5: forest_managed
• Layer 7: other

All three of these land-use classes show some pixels with big differences among the three scenarios, from 2025 onwards. Many of the differences are nonintuitive; but none appears to be in the Andes.

Because the solution does not appear to lie within those three land-use classes, the same plots are also then used for the other 9 classes, in numerical order (but they come after the three listed above).

```{r compare-netCDFs, echo=FALSE, cache=TRUE}

bluered.plot <- function(raster1, raster2, main="", ylim=c(-90,90), xlim=c(-180,180)){
  # Slightly kludgy but plots differences within the range -1 to +1, with zero being white
  mindiff <- round(cellStats(raster1-raster2, "min"), 2)
  maxdiff <- round(cellStats(raster1-raster2, "max"), 2)
  if (maxdiff - mindiff < 0.01){
    print(paste(main, year[i],": differences too small, between", mindiff, "and", maxdiff))
  }else{
    mycols <- bluered(200)
    tooblue <- 100 + (mindiff * 100)
    toored <- (1 -maxdiff) * 100
    mycols <- mycols[1:(200-toored)]
    mycols <- mycols[-c(1:tooblue)]
    plot(raster1 - raster2, col=mycols, ylim=ylim, xlim=xlim, main=main, sub=paste("Difference range is", mindiff, "to", maxdiff))
    map(add=TRUE, col="lightgrey", interior=FALSE)
  }
}

stem <- "data/IAM outputs for 30 by 30 scenarios/IMAGE"

file_BID <- "BendingTheCurve30by30-LCproj-IMAGE-RCPref_SSP2_BID-06Feb2020_FinalMask.nc"
file_CRJ <- "BendingTheCurve30by30-LCproj-IMAGE-RCPref_SSP2_CRJ-06Feb2020_FinalMask.nc"
file_HPR <- "BendingTheCurve30by30-LCproj-IMAGE-RCPref_SSP2_HPR-06Feb2020_FinalMask.nc"

nc_BID <- nc_open(file.path(stem, file_BID))
print(strsplit(nc_BID$dim$lc_class$units, "/"))
year <- nc_BID$dim$time$vals
print(year)

# forest_unmanaged
fu_BID <- brick(file.path(stem, file_BID), nl=nl, varname="LC_area_share", level=4)
fu_CRJ <- brick(file.path(stem, file_CRJ), nl=nl, varname="LC_area_share", level=4)
fu_HPR <- brick(file.path(stem, file_HPR), nl=nl, varname="LC_area_share", level=4)

fu_BIDminusCRJ <- fu_BID - fu_CRJ
fu_CRJminusHPR <- fu_CRJ - fu_HPR

par(mfrow=c(1,2))

for (i in 1:5){
  #Loop through first 5 layers of each stack
  bluered.plot(subset(fu_BID,i), subset(fu_CRJ, i), main="BID - CRJ: forest_unmanaged", xlim=c(-85,-50), ylim=c(-15, 10))
  bluered.plot(subset(fu_CRJ,i), subset(fu_HPR, i), main="CRJ - HPR: forest_unmanaged", xlim=c(-85,-50), ylim=c(-15, 10))
}


# forest_managed
fm_BID <- brick(file.path(stem, file_BID), nl=nl, varname="LC_area_share", level=5)
fm_CRJ <- brick(file.path(stem, file_CRJ), nl=nl, varname="LC_area_share", level=5)
fm_HPR <- brick(file.path(stem, file_HPR), nl=nl, varname="LC_area_share", level=5)

fm_BIDminusCRJ <- fm_BID - fm_CRJ
fm_CRJminusHPR <- fm_CRJ - fm_HPR

par(mfrow=c(1,2))

for (i in 1:5){
  #Loop through first 5 layers of each stack
  bluered.plot(subset(fm_BID,i), subset(fm_CRJ, i), main="BID - CRJ: forest_managed", xlim=c(-85,-50), ylim=c(-15, 10))
  bluered.plot(subset(fm_CRJ,i), subset(fm_HPR, i), main="CRJ - HPR: forest_managed", xlim=c(-85,-50), ylim=c(-15, 10))
}


# other
o_BID <- brick(file.path(stem, file_BID), nl=nl, varname="LC_area_share", level=7)
o_CRJ <- brick(file.path(stem, file_CRJ), nl=nl, varname="LC_area_share", level=7)
o_HPR <- brick(file.path(stem, file_HPR), nl=nl, varname="LC_area_share", level=7)

o_BIDminusCRJ <- o_BID - o_CRJ
o_CRJminusHPR <- o_CRJ - o_HPR

par(mfrow=c(1,2))

for (i in 1:5){
  #Loop through first 5 layers of each stack
  bluered.plot(subset(o_BID,i), subset(o_CRJ, i), main="BID - CRJ: other", xlim=c(-85,-50), ylim=c(-15, 10))
  bluered.plot(subset(o_CRJ,i), subset(o_HPR, i), main="CRJ - HPR: other", xlim=c(-85,-50), ylim=c(-15, 10))
}


# cropland_other
co_BID <- brick(file.path(stem, file_BID), nl=nl, varname="LC_area_share", level=1)
co_CRJ <- brick(file.path(stem, file_CRJ), nl=nl, varname="LC_area_share", level=1)
co_HPR <- brick(file.path(stem, file_HPR), nl=nl, varname="LC_area_share", level=1)

par(mfrow=c(1,2))

for (i in 1:5){
  #Loop through first 5 layers of each stack
  bluered.plot(subset(co_BID,i), subset(co_CRJ, i), main="BID - CRJ: cropland_other", xlim=c(-85,-50), ylim=c(-15, 10))
  bluered.plot(subset(co_CRJ,i), subset(co_HPR, i), main="CRJ - HPR: cropland_other", xlim=c(-85,-50), ylim=c(-15, 10))
}

# cropland_2Gbioen
c2_BID <- brick(file.path(stem, file_BID), nl=nl, varname="LC_area_share", level=2)
c2_CRJ <- brick(file.path(stem, file_CRJ), nl=nl, varname="LC_area_share", level=2)
c2_HPR <- brick(file.path(stem, file_HPR), nl=nl, varname="LC_area_share", level=2)

for (i in 1:5){
  #Loop through first 5 layers of each stack
  bluered.plot(subset(c2_BID,i), subset(c2_CRJ, i), main="BID - CRJ: cropland_2Gbioen", xlim=c(-85,-50), ylim=c(-15, 10))
  bluered.plot(subset(c2_CRJ,i), subset(c2_HPR, i), main="CRJ - HPR: cropland_2Gbioen", xlim=c(-85,-50), ylim=c(-15, 10))
}

# grassland
g_BID <- brick(file.path(stem, file_BID), nl=nl, varname="LC_area_share", level=3)
g_CRJ <- brick(file.path(stem, file_CRJ), nl=nl, varname="LC_area_share", level=3)
g_HPR <- brick(file.path(stem, file_HPR), nl=nl, varname="LC_area_share", level=3)

for (i in 1:5){
  #Loop through first 5 layers of each stack
  bluered.plot(subset(g_BID,i), subset(g_CRJ, i), main="BID - CRJ: grassland", xlim=c(-85,-50), ylim=c(-15, 10))
  bluered.plot(subset(g_CRJ,i), subset(g_HPR, i), main="CRJ - HPR: grassland", xlim=c(-85,-50), ylim=c(-15, 10))
}

# restored
r_BID <- brick(file.path(stem, file_BID), nl=nl, varname="LC_area_share", level=6)
r_CRJ <- brick(file.path(stem, file_CRJ), nl=nl, varname="LC_area_share", level=6)
r_HPR <- brick(file.path(stem, file_HPR), nl=nl, varname="LC_area_share", level=6)

for (i in 1:5){
  #Loop through first 5 layers of each stack
  bluered.plot(subset(r_BID,i), subset(r_CRJ, i), main="BID - CRJ: restored", xlim=c(-85,-50), ylim=c(-15, 10))
  bluered.plot(subset(r_CRJ,i), subset(r_HPR, i), main="CRJ - HPR: restored", xlim=c(-85,-50), ylim=c(-15, 10))
}

# built-up
b_BID <- brick(file.path(stem, file_BID), nl=nl, varname="LC_area_share", level=8)
b_CRJ <- brick(file.path(stem, file_CRJ), nl=nl, varname="LC_area_share", level=8)
b_HPR <- brick(file.path(stem, file_HPR), nl=nl, varname="LC_area_share", level=8)

for (i in 1:5){
  #Loop through first 5 layers of each stack
  bluered.plot(subset(b_BID,i), subset(b_CRJ, i), main="BID - CRJ: built-up", xlim=c(-85,-50), ylim=c(-15, 10))
  bluered.plot(subset(b_CRJ,i), subset(b_HPR, i), main="CRJ - HPR: built-up", xlim=c(-85,-50), ylim=c(-15, 10))
}

# abn_cropland_other
aco_BID <- brick(file.path(stem, file_BID), nl=nl, varname="LC_area_share", level=9)
aco_CRJ <- brick(file.path(stem, file_CRJ), nl=nl, varname="LC_area_share", level=9)
aco_HPR <- brick(file.path(stem, file_HPR), nl=nl, varname="LC_area_share", level=9)

for (i in 1:5){
  #Loop through first 5 layers of each stack
  bluered.plot(subset(aco_BID,i), subset(aco_CRJ, i), main="BID - CRJ: abn_cropland_other", xlim=c(-85,-50), ylim=c(-15, 10))
  bluered.plot(subset(aco_CRJ,i), subset(aco_HPR, i), main="CRJ - HPR: abn_cropland_other", xlim=c(-85,-50), ylim=c(-15, 10))
}

# abn_cropland_2Gbioen
ac2_BID <- brick(file.path(stem, file_BID), nl=nl, varname="LC_area_share", level=10)
ac2_CRJ <- brick(file.path(stem, file_CRJ), nl=nl, varname="LC_area_share", level=10)
ac2_HPR <- brick(file.path(stem, file_HPR), nl=nl, varname="LC_area_share", level=10)

for (i in 1:5){
  #Loop through first 5 layers of each stack
  bluered.plot(subset(ac2_BID,i), subset(ac2_CRJ, i), main="BID - CRJ: abn_cropland_2Gbioen", xlim=c(-85,-50), ylim=c(-15, 10))
  bluered.plot(subset(ac2_CRJ,i), subset(ac2_HPR, i), main="CRJ - HPR: abn_cropland_2Gbioen", xlim=c(-85,-50), ylim=c(-15, 10))
}

# abn_grassland
ag_BID <- brick(file.path(stem, file_BID), nl=nl, varname="LC_area_share", level=11)
ag_CRJ <- brick(file.path(stem, file_CRJ), nl=nl, varname="LC_area_share", level=11)
ag_HPR <- brick(file.path(stem, file_HPR), nl=nl, varname="LC_area_share", level=11)

for (i in 1:5){
  #Loop through first 5 layers of each stack
  bluered.plot(subset(ag_BID,i), subset(ag_CRJ, i), main="BID - CRJ: abn_grassland", xlim=c(-85,-50), ylim=c(-15, 10))
  bluered.plot(subset(ag_CRJ,i), subset(ag_HPR, i), main="CRJ - HPR: abn_grassland", xlim=c(-85,-50), ylim=c(-15, 10))
}

# abn_forest_managed
afm_BID <- brick(file.path(stem, file_BID), nl=nl, varname="LC_area_share", level=12)
afm_CRJ <- brick(file.path(stem, file_CRJ), nl=nl, varname="LC_area_share", level=12)
afm_HPR <- brick(file.path(stem, file_HPR), nl=nl, varname="LC_area_share", level=12)

for (i in 1:5){
  #Loop through first 5 layers of each stack
  bluered.plot(subset(afm_BID,i), subset(afm_CRJ, i), main="BID - CRJ: abn_forest_managed", xlim=c(-85,-50), ylim=c(-15, 10))
  bluered.plot(subset(afm_CRJ,i), subset(afm_HPR, i), main="CRJ - HPR: abn_forest_managed", xlim=c(-85,-50), ylim=c(-15, 10))
}

```
## Next steps

I wonder whether it's possible that one of the scenarios didn't get one of the input layers incorporated into it in the first place, or multiplied by the 'wrong' cell area raster, perhaps explaining how the ENA shows a 'deficit' for the Andes despite there not being anything evident in the netCDFs. That at least seems worth looking into next.

Looking into the BID IMAGE data, there's a huge disjunction in the managed forest between 2020 and 2025, with 100 Mha being lost. If that's being lost in the areas where endemic richness then fell, I have a likely explanation for what's gone on.

Looking more closely, there's a lot of managed forest that becomes abandoned managed forest, and which then gets a \emph{lower} BII value as a result. I have therefore edited convert_iam_data.R so that the BII coefficient for abandoned managed forest is at least as high as that for managed forest. However, this hasn't made any real difference (the global ENA through time is barely different, and it does not remove the big differences in 2030 ENA between CRJ and HPR).

What is causing the difference in the Andes in 2030?


```{r recalculate-ENA-and-compare, echo=FALSE, cache=TRUE}
source("convert_iam_data.R")
print("Trying now with a constraint on BII in adandoned managed forest...")
in_stem <- "data/IAM outputs for 30 by 30 scenarios/IMAGE"
which_bii <- "BII_humdom0"
ena_stem <- "output/checking"
pdf_stem <- "reports/checking"
BID_file <- "BendingTheCurve30by30-LCproj-IMAGE-RCPref_SSP2_BID-06Feb2020_FinalMask.nc"
new_ena_BID <- convert_iam_data(file.path(in_stem, BID_file), which_bii, file.path(ena_stem, paste("BII", which_bii, "BID_ENA.nc", sep="_")), file.path(pdf_stem, paste("BII", which_bii, "BID_ENA.pdf", sep="_")), layers=c(1:19))

CRJ_file <- "BendingTheCurve30by30-LCproj-IMAGE-RCPref_SSP2_CRJ-06Feb2020_FinalMask.nc"
new_ena_CRJ <- convert_iam_data(file.path(in_stem, CRJ_file), which_bii, file.path(ena_stem, paste("BII", which_bii, "CRJ_ENA.nc", sep="_")), file.path(pdf_stem, paste("BII", which_bii, "CRJ_ENA.pdf", sep="_")), layers=c(1:19))

HPR_file <- "BendingTheCurve30by30-LCproj-IMAGE-RCPref_SSP2_HPR-06Feb2020_FinalMask.nc"
new_ena_HPR <- convert_iam_data(file.path(in_stem, HPR_file), which_bii, file.path(ena_stem, paste("BII", which_bii, "HPR_ENA.nc", sep="_")), file.path(pdf_stem, paste("BII", which_bii, "HPR_ENA.pdf", sep="_")), layers=c(1:19))

plot(subset(new_ena_CRJ,5) - subset(new_ena_HPR, 5)) # Global
plot(subset(new_ena_CRJ,5) - subset(new_ena_HPR, 5), xlim=c(-85,-50), ylim=c(-15, 10)) # Zoom in on Andes

```

## Mapping the loss of managed forest between 2020 and 2025
plot(subset(new_ena_CRJ,5) - subset(new_ena_HPR, 5))
If the loss of managed forest were concentrated in, e.g., the Andes, it might be relevant; but it's not.

```{r mapping_forest_managed, echo=FALSE, cache=TRUE}
for (i in 1:4){
  bluered.plot(subset(fm_BID, i), subset(fm_BID, (i+1)), main=paste("BID forest_managed:", year[i], "-", year[i+1]))
}
```