---
title: "Using the IMAGE historical data"
author: "Andy Purvis"
date: "18/02/2021"
output:
  html_document: 
    df_print: paged
    code_folding: hide
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
  pdf_document:
    df_print: kable
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Using the IMAGE historical data

THIS RUN IS A SENSITIVITY ANALYSIS USING SQRT(RSR)
David LeClere has shared with me a 30-by-30-compatible historical land-use netCDF going back to 1970. This is appealing for two main reasons. First, IMAGE provides a better match to the LUH2 data in 2010 than do the other IAMs (MAgPIE, AIM and GLOBIOM). Second, 1970 is a full half-life before the present, de-emphasizing artifacts caused by having to harmonize incongruent land-use trajectories.

This code has the following objectives:

1. Read in and examine the IMAGE data;

2. Compare IMAGE with LUH2 for each time step from 1970 to 2015, hoping that 1970 will not be very much worse than 2015; and using the range-size rarity layer when making these comparisons;

3. Harmonizing the effective natural area (ENA) data from LUH2 with those from IMAGE, ideally using 1970 as the year on which to harmonize;

4. Run the relaxation to infer the global trajectory of extinctions of endemic species dependent on natural habitat, from 850 to 2020; and

5. Run the relaxation using the ENA estimates for 2020-2100 under each of the seven IMAGE scenarios, to infer the future global extinction trajectories of endemic habitat-specialist species.

## Reproducibility

The checkpoint package is used to aid reproducibility.

```{r, load, echo=FALSE, cache=FALSE}
library(checkpoint)
checkpoint("2021-02-16")

start.time <- Sys.time()
library(raster)
library(scales)
library(abind)
library(gplots)
library(colorspace)
library(maps)
library(ncdf4) # package for netcdf manipulation, suggested by https://rpubs.com/boyerag/297592
library(rgdal) # package for geospatial analysis, suggested by https://rpubs.com/boyerag/297592
library(ggplot2) # package for plotting, suggested by https://rpubs.com/boyerag/297592
source("00_functions.R")
source("convert_iam_data.R")
```
## Read in range-size rarity to assess how badly mismatches between IMAGE and LUH2 might matter

Difficulties harmonizing the data on effective natural area matter more in regions with considerable range-size rarity, as these are the places where extinction debt could be artificially inflated by mismatches. The range-size rarity layer is also used in the dynamic species-area modelling, as a proxy for the geographic pattern of diversity of habitat-specialist endemics in the year 2015.

```{r, range-size-rarity, echo=FALSE, cache=TRUE}
rsr <- raster("output/rsr2.nc") # Generated as part of De Palma et al. (in prep.), as the sum of the 44 grid cells per 2-degree cell

rsr_rescaled <- rsr/cellStats(rsr, "mean")
plot(rsr_rescaled, main="Relative range-size rarity per 2-degree grid cell in 2015")
map(add=TRUE)
rsr_array <- as.array(rsr_rescaled) #Convert to array as S_t is an array
```

## Reading the file in and examining the metadata

The IMAGE historical data are on an 0.5$^\circ$ grid, and there are four years -- 1970, 1980, 1990 and 2000. There are the same 12 land-use classes as used by all the IAMs in the 30 by 30 project. As well as the stack containing the area shares of each land-use class, there is a single raster of pixel (cell) area. The structure is therefore not exactly the same as the IAM data used previously: there are fewer (as well as different) years, but also the internal organisation is sufficiently different that the years with data could not be identified using the same code as for those files.

```{r read-and-examine, echo=FALSE}
image <- brick("data/ForAndy/BendingTheCurveFT-LCproj-IMAGE-history-R2_wabn_31Jan2020_FinalMask.nc", level=3)
print(image)
cat("\nYears with land-use data:\n")
print(unlist(image@z)) # Which years
```

## Estimating ENA for each of the four layers (no maps produced)

The IMAGE file is read in, and the BII_humdom0 set of BII coefficients used to translate the mix of land-use classes into estimates of each grid cell's effective natural area (ENA).

```{r, get-ENA-for-IMAGE, echo=FALSE, cache=TRUE}
which_bii <- "BII_humdom0"
iam_file <- "data/ForAndy/BendingTheCurveFT-LCproj-IMAGE-history-R2_wabn_31Jan2020_FinalMask.nc"
ena_file <- "output/ena/ENA-IMAGE-1970-2000.nc"
pdf_file <- "reports/Converting IAMs to ENA/ENA-IMAGE-1970-2000.pdf"
ena_image_1970 <- convert_iam_data(iam_file, which_bii, ena_file, pdf_file, yb=1970, ye=2000, ys=10)
cell_areas <- aggregate(raster(iam_file, varname = "pixel_area"), fact=4, fun=sum)

```

## Compare IMAGE at 1970, 1980, 1990 and 2000 with corresponding LUH2 data

The results of this comparison are extremely promising First, the agreement is best in 1970, which is the preferred year on which to harmonize for conceptual reasons as well. Second, the agreement is improved when cells are weighted by their range-size rarity, meaning that the disagreements matter less than they might have done.

```{r, compare-with-LUH2, echo=FALSE, cache=TRUE}
ena <- stack("output/ena/ENA-IMAGE-1970-2000.nc")
luh2 <- raster("output/ENA_LUH2_1970.nc")
luh2 <- addLayer(luh2, raster("output/ENA_LUH2_1980.nc"))
luh2 <- addLayer(luh2, raster("output/ENA_LUH2_1990.nc"))
luh2 <- addLayer(luh2, raster("output/ENA_LUH2_2000.nc"))

years <- c(1970,1980,1990,2000)
for (i in 1:4){
  plot(subset(luh2, i) - subset(ena, i), main = paste("LUH2 - IMAGE:", years[i]))
  map(add=TRUE)
  plot(subset(luh2, i), subset(ena, i), ylab=paste("IMAGE", years[i]), xlab=paste("LUH2", years[i]))
  abline(a=0,b=1,col="red")
  lv <- getValues(subset(luh2,i))
  iv <- getValues(subset(ena, i))
  li <- !is.na(lv*iv)
  text(0, 4.5, paste("r =", round(cor(lv[li], iv[li]), 3)), adj=0)
  
  m1 <- lm(iv[li] ~ lv[li], weights=getValues(rsr_rescaled)[li])
  text(0, 4, paste("weighted r =", round(sqrt(summary(m1)$r.squared), 3)), adj=0)

}
```


## Produce annual LUH2 data and rescale to match IMAGE data

LUH2 data cover the years 850-2015 in five-year time steps. However, to ensure that species cannot be resurrected, even this level of granularity is not sufficient; the ENA estimates are therefore used as the basis for interpolation to produce annual ENA maps (in the form of a 3-dimensional array).

The source file read in to produce the historic_annual_area array has been somewhat hard-coded to accommodate the unique case of the IMAGE historical estimates; this decision was taken for speed of getting the IMAGE results, but it also has the consequence that a new version of the source file will be needed for the other IAMs, slowing down their results. Harmonization used the 1970 IMAGE data.

The first plot shows the global trajectory of effective natural area over time. The second shows that there is a discontinuity in the rate of change in ENA at 1970, but it still represents a very small fraction of the overall ENA value. Nonetheless, it could provide problems when other IAMs are considered, for two reasons. First, the other IAMs fit LUH2 less well, meaning the discontinuity is likely to be larger. Second, harmonization will have to be done for the year 2010, meaning there will have been much less time for relaxation to smooth away artifacts.

```{r, rescale-LUH2, echo=FALSE, cache=TRUE}
if (file.exists("output/historic_annual_area.Rds")){
  historic_annual_area <- readRDS("output/historic_annual_area.Rds")
}else{
  source("01_produce_historical_area.R")
}

# Summarise landscape change
global_effective_area <- apply(historic_annual_area, MARGIN=3, sum, na.rm=TRUE)
plot(global_effective_area ~ c(850:2020), ylab="Global natural habitat", xlab="Year", type="l",
     main="Global effective area from LUH2 harmonized with IMAGE in 1970", sub="Note rise after Black Death")
abline(v=1970, col="red", lty=2)

# Check for a discontinuity in rate of change - i.e., in the sum of absolute year-to-year changes across grid cells
turnover <- rep(NA, 2020-850)
for (i in 1:(2020-850)){
  diff <- abs(historic_annual_area[,,i+1] - historic_annual_area[,,i])
  turnover[i] <- sum(diff, na.rm=TRUE)
}
plot(turnover ~ c(850:2019), type="l", ylab="Gross land-use change",
     main = "Land-use change between successive years") 
abline(v=1970, col="red", lty=2)

```

## Do relaxation from 850 to 2020

The dynamic species-area relationship models how extinctions accumulate after the loss of effective natural area, using parameters from Wearn et al. (2012). For the historical land-use data (i.e., LUH2 plus historical IMAGE data plus the 2015 and 2020 layers from one of the IMAGE scenarios), this relaxation is shown in two ways. The graph shows the percentage losses of both ENA and endemic species, while the map shows where in the world the extinctions are inferred as having taken place.


```{r, relax_to_2020, echo=FALSE, warning=FALSE, cache=TRUE}
print(paste(Sys.time(), "Analysing relaxation from 850 to 2020 ..."))
this.k <- 0.0122
this.z <- 0.269
relaxation_history <- relax(natarea=historic_annual_area, years=c(850:2020), 
                            wanted=c(850:2020), rsr_2015=rsr_array, k=this.k, z=this.z)
S_t_hist <- relaxation_history$S_t
c_hist <- relaxation_history$c_array
endemics_hist <- apply(S_t_hist, MARGIN=3, sum, na.rm=TRUE)
haa <- apply(historic_annual_area, MARGIN=3, sum, na.rm=TRUE)
# Plot, map and totals
percentage_plot(A=historic_annual_area, S=S_t_hist, year=c(850:2020), main_stem="Historical")
abline(v=1970, col="red", lty=2)
extinctions_map(S=S_t_hist, year=c(850:2020), between=c(1850,2020), main_stem="Historical")
history_totals <- get_totals(S = endemics_hist, year = c(850:2020))
cat("850-2020: extinction summary\n")
print(history_totals)
print(Sys.time())
```
## Loop through IMAGE scenarios
For each IMAGE scenario, the effective natural area (ENA) layers from 2010 to 2100 in 5-year steps are first converted to an array of annual (interpolated) ENA. This annual ENA array is then used to drive relaxation. The first ten years should be identical to the last ten years of the relaxation over the historical period (as a sense-check), with the scenarios beginning to diverge only after 2020.

Each scenario's results are shown in three ways: a graph of the losses of ENA and endemic species from 2020-2100, expressed as percentages of the levels in 2020; a map of extinctions projected to take place between 2020 and 2100; and a summary of the relative numbers of endemics remaining in different selected years.

```{r, relax_to_2100, echo=FALSE, warning=FALSE, cache=TRUE}
image_files <- c("ENA-IMAGE-RCPref_SSP2_BID-06Feb2020_FinalMask.nc",
                 "ENA-IMAGE-RCPref_SSP2_CRJ-06Feb2020_FinalMask.nc",
                 "ENA-IMAGE-RCPref_SSP2_GDN-06Feb2020_FinalMask.nc",
                 "ENA-IMAGE-RCPref_SSP2_HPR-06Feb2020_FinalMask.nc",
                 "ENA-IMAGE-RCPref_SSP2_NRS-06Feb2020_FinalMask.nc",
                 "ENA-IMAGE-RCPref_SSP2_THC-06Feb2020_FinalMask.nc",
                 "ENA-IMAGE-RCPref_SSP2_WAC-06Feb2020_FinalMask.nc")
scenarios <- c("BID", "CRJ", "GDN", "HPR", "NRS", "THC", "WAC")
shortnames <- gsub("ENA-IMAGE-RCPref_SSP2_", "aslu", image_files)
pre2010 <- historic_annual_area[,,c(1:1160)] # 850 to 2009

for (i in 1:length(image_files)){
  print(paste(Sys.time(), "Processing", image_files[i], "..."))
  slu <- stack(file.path("output/ena", image_files[i]))
  names(slu) <- seq(2010, 2100, by = 5)
  aslu <- interpolateTemporal(slu, xin=seq(2010, 2100, by=5), xout=seq(2010, 2100, by=1), 
                              outdir="figs", prefix=paste("IMAGE_", scenarios[i], sep=""), progress=TRUE, 
                                   writechange=FALSE, returnstack=TRUE, overwrite=TRUE)
  if (require(ncdf4)) {	
   rnc <- writeRaster(aslu, filename=file.path("output/aslu", paste("IMAGE_", shortnames[i], sep="")), format="CDF", overwrite=TRUE) 
  }

  # convert aslu to an array and save it
  aslu_array <- abind(pre2010, as.array(aslu))
  saveRDS(aslu_array, file.path("output/arrays/", paste(image_files[i], ".Rds", sep="")))
  area_aslu  <- apply(aslu_array, MARGIN=3, sum, na.rm=TRUE)

  # relax using aslu array
  relaxation_scenario <- relax(natarea=aslu_array, years=c(850:2100), wanted=c(2010:2100), rsr_2015=rsr_array, k=this.k, z=this.z)
  S_t_scen <- relaxation_scenario$S_t
  saveRDS(S_t_scen, file.path("output/arrays/", paste(shortnames[i], ".Rds", sep="")))

  # Calculate numbers of endemic species and area, do percentage plots, and save time series
  endemics_scen <- apply(S_t_scen, MARGIN=3, sum, na.rm=TRUE)
  saveRDS(endemics_scen, file.path("output/endemics", paste(shortnames[i], ".Rds", sep="")))

  percentage_plot(A=aslu_array[,,c(1171:1251)], S=S_t_scen[,,c(11:91)], year=c(2020:2100), main_stem=shortnames[i])
  extinctions_map(S=S_t_scen, year=c(2010:2100), between=c(2020, 2100), main_stem=shortnames[i])
  scen_totals <- get_totals(S = endemics_scen, year = c(2020:2100))
  cat(paste(shortnames[i], "2020-2100: extinction summary\n"))
  print(scen_totals)
}


```

## Put results from scenarios into a data frame and plot them

The results are most naturally expressed as cumulative numbers of extinctions over time, on a relative scale. The two scenarios that prevent most extinctions are GDN and (surprisingly?) HPR, with NRS (unsurprisingly) last. The best two prevent over 15% of the extinction projected under the baseline scenario (NRS). The main reason these differences are not greater is that the scenarios do not include restoration, which can avoid repayment of current extinction debt. 

```{r, harvest-results, echo=FALSE, warning=FALSE, cache=TRUE}
image_projections <- data.frame(year = c(2010:2100))
image_projections$BID <- readRDS(file.path("output/endemics","asluBID-06Feb2020_FinalMask.nc.Rds"))
image_projections$CRJ <- readRDS(file.path("output/endemics","asluCRJ-06Feb2020_FinalMask.nc.Rds"))
image_projections$GDN <- readRDS(file.path("output/endemics","asluGDN-06Feb2020_FinalMask.nc.Rds"))
image_projections$HPR <- readRDS(file.path("output/endemics","asluHPR-06Feb2020_FinalMask.nc.Rds"))
image_projections$NRS <- readRDS(file.path("output/endemics","asluNRS-06Feb2020_FinalMask.nc.Rds"))
image_projections$THC <- readRDS(file.path("output/endemics","asluTHC-06Feb2020_FinalMask.nc.Rds"))
image_projections$WAC <- readRDS(file.path("output/endemics","asluWAC-06Feb2020_FinalMask.nc.Rds"))

image_projections <- subset(image_projections, year >= 2020)

for(i in 2:8){
  image_projections[,i] <- image_projections[1,i] - image_projections[,i]
} 

toll <- max(image_projections[,c(2:8)])
image_projections[,c(2:8)] <- 100*image_projections[, c(2:8)]/toll

plot(BID ~ year, data=image_projections, col="red", type="l", ylim=c(0,100),
     ylab="Extinctions since 2020 (relative)", xlab="Year", cex.lab=1.3)
text(2020, 95, adj=0, "Key to scenarios:")
text(2020, 90, adj=0, "NRS", col="blue")
text(2020, 85, adj=0, "THC", col="darkblue")
text(2020, 80, adj=0, "CRJ", col="orange")
text(2020, 75, adj=0, "WAC", col="violet")
text(2020, 70, adj=0, "BID", col="red")
text(2020, 65, adj=0, "GDN", col="yellow")
text(2020, 60, adj=0, "HPR", col="green")

lines(CRJ ~ year, data=image_projections, col="orange", type="l")
lines(GDN ~ year, data=image_projections, col="yellow", type="l")
lines(HPR ~ year, data=image_projections, col="green", type="l")
lines(NRS ~ year, data=image_projections, col="blue", type="l")
lines(THC ~ year, data=image_projections, col="darkblue", type="l")
lines(WAC ~ year, data=image_projections, col="violet", type="l")


saveRDS(image_projections, "output/endemics/image_projections.Rds")
```

## Internal consistency check

The period 2010-2020 has been modelled both as part of the history and as part of the scenarios. The two estimates should be the same.


```{r, check-internal-consistency, echo=FALSE, cache=TRUE}
hh <- tail(endemics_hist, 11)
ss <- endemics_scen[1:11]
plot(hh, ss)
abline(a=0,b=1)
```
