---
title: "Using_AIM_data"
author: "Andy Purvis"
date: "28/02/2021"
output:
  html_document: 
    df_print: paged
    code_folding: hide
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
  pdf_document:
    df_print: kable
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Using the IAM data

The IMAGE data had the best fit to LUH2, but an odd pattern of land conversion in the years immediately following 2020 gave some highly non-intuitive results. Of the other three IAMs, AIM has the best fit to LUH2. This markdown therefore harmonizes the AIM scenario data to LUH2 and performs the dynamic species-area analysis.

## Reproducibility

The checkpoint package is used to aid reproducibility.

```{r, load, echo=FALSE, cache=FALSE}
library(checkpoint)
checkpoint("2021-02-16")

start.time <- Sys.time()
library(raster)
library(scales)
library(abind)
library(gplots)
library(colorspace)
library(maps)
library(ncdf4) # package for netcdf manipulation, suggested by https://rpubs.com/boyerag/297592
library(rgdal) # package for geospatial analysis, suggested by https://rpubs.com/boyerag/297592
library(ggplot2) # package for plotting, suggested by https://rpubs.com/boyerag/297592
source("00_functions.R")
source("convert_iam_data.R")
source("harmonize_with_luh2_2010.R")

```

## Read in range-size rarity to assess how badly mismatches between AIM and LUH2 might matter

Difficulties harmonizing the data on effective natural area matter more in regions with considerable range-size rarity, as these are the places where extinction debt could be artificially inflated by mismatches. The range-size rarity layer is also used in the dynamic species-area modelling, as a proxy for the geographic pattern of diversity of habitat-specialist endemics in the year 2015.

```{r, range-size-rarity, echo=FALSE, cache=TRUE}
rsr <- raster("output/rsr.nc") # Generated as part of De Palma et al. (in prep.), as the sum of the 44 grid cells per 2-degree cell
rsr_rescaled <- rsr/cellStats(rsr, "mean")
plot(rsr_rescaled, main="Relative range-size rarity per 2-degree grid cell in 2015")
map(add=TRUE)
rsr_array <- as.array(rsr_rescaled) #Convert to array as S_t is an array
```
## Compare AIM ENA estimates for 2010 with corresponding LUH2 data and rescale LUH2

The across-cells correlation between ENA estimates from LUH2 and AIM is 0.86; weighting by range-size rarity reduces this to 0.827. The map shows that some Andean cells are almost entirely natural land according to LUh2 and almost entirely converted land according to AIM.

```{r, compare-with-LUH2, echo=FALSE, cache=TRUE}
cell_area_raster <- raster("data/cell-area.tif")
ena <- stack("output/ena/ENA-AIM-RCPref_SSP2_BID-12Feb2020_FinalMask.nc")
luh2_2010 <- raster("output/ENA_LUH2_2010.nc")

plot(luh2_2010 - subset(ena, 1), main = "LUH2 - AIM: 2010")
map(add=TRUE)
plot(luh2_2010, subset(ena, 1), ylab="AIM 2010", xlab="LUH2 2010")
abline(a=0,b=1,col="red")
lv <- getValues(luh2_2010)
iv <- getValues(subset(ena, 1))
li <- !is.na(lv*iv)
text(0, 4.5, paste("r =", round(cor(lv[li], iv[li]), 3)), adj=0)
  
m1 <- lm(iv[li] ~ lv[li], weights=getValues(rsr_rescaled)[li])
text(0, 4, paste("weighted r =", round(sqrt(summary(m1)$r.squared), 3)), adj=0)

```
## Produce annual LUH2 data and rescale to match IMAGE data

LUH2 data cover the years 850-2015 in five-year time steps. However, to ensure that species cannot be resurrected, even this level of granularity is not sufficient; the ENA estimates are therefore used as the basis for interpolation to produce annual ENA maps (in the form of a 3-dimensional array).

The source file read in to produce the historic_annual_area array has been somewhat hard-coded to accommodate the unique case of the IMAGE historical estimates; this decision was taken for speed of getting the IMAGE results, but it also has the consequence that a new version of the source file will be needed for the other IAMs, slowing down their results. Harmonization used the 1970 IMAGE data.

The first plot shows the global trajectory of effective natural area over time. The second shows that there is a discontinuity in the rate of change in ENA at 1970, but it still represents a very small fraction of the overall ENA value. Nonetheless, it could provide problems when other IAMs are considered, for two reasons. First, the other IAMs fit LUH2 less well, meaning the discontinuity is likely to be larger. Second, harmonization will have to be done for the year 2010, meaning there will have been much less time for relaxation to smooth away artifacts.

```{r, rescale-LUH2, echo=FALSE, cache=TRUE}
historic_annual_area_aim <- 
  harmonize_with_luh2_2010("output/ena/ENA-AIM-RCPref_SSP2_BID-12Feb2020_FinalMask.nc", "AIM")

# Summarise landscape change
global_effective_area_aim <- apply(historic_annual_area_aim, MARGIN=3, sum, na.rm=TRUE)
plot(global_effective_area_aim ~ c(850:2015), ylab="Global natural habitat", xlab="Year", type="l",
     main="Global effective area from LUH2 harmonized with AIM in 2010")
abline(v=2010, col="red", lty=2)

# Check for a discontinuity in rate of change - i.e., in the sum of absolute year-to-year changes across grid cells
turnover <- rep(NA, 2015-850)
for (i in 1:(2015-850)){
  diff <- abs(historic_annual_area_aim[,,i+1] - historic_annual_area_aim[,,i])
  turnover[i] <- sum(diff, na.rm=TRUE)
}
plot(turnover ~ c(850:2014), type="l", ylab="Gross land-use change",
     main = "AIM: land-use change between successive years") 
abline(v=2010, col="red", lty=2)

```



## Do relaxation from 850 to 2020

The dynamic species-area relationship models how extinctions accumulate after the loss of effective natural area, using parameters from Wearn et al. (2012). For the historical land-use data (i.e., LUH2 plus the 2015 and 2020 layers from one of the AIM scenarios), this relaxation is shown in two ways. The graph shows the percentage losses of both ENA and endemic species, while the map shows where in the world the extinctions are inferred as having taken place.


```{r, relax_to_2010, echo=FALSE, warning=FALSE, cache=TRUE}
print(paste(Sys.time(), "Analysing relaxation from 850 to 2015 ..."))
this.k <- 0.0122
this.z <- 0.269
relaxation_history_aim <- relax(natarea=historic_annual_area_aim, years=c(850:2015), 
                            wanted=c(850:2015), rsr_2015=rsr_array, k=this.k, z=this.z)
S_t_hist_aim <- relaxation_history_aim$S_t
c_hist_aim <- relaxation_history_aim$c_array
endemics_hist_aim <- apply(S_t_hist_aim, MARGIN=3, sum, na.rm=TRUE)
haa_aim <- apply(historic_annual_area_aim, MARGIN=3, sum, na.rm=TRUE)
# Plot, map and totals
percentage_plot(A=historic_annual_area_aim, S=S_t_hist_aim, year=c(850:2015), main_stem="Historical")
abline(v=2010, col="red", lty=2)
extinctions_map(S=S_t_hist_aim, year=c(850:2015), between=c(1850,2015), main_stem="Historical: AIM")
history_totals_aim <- get_totals(S = endemics_hist_aim, year = c(850:2010))
cat("850-2010: extinction summary\n")
print(history_totals_aim)
print(Sys.time())
```

## Loop through AIM scenarios
For each AIM scenario, the effective natural area (ENA) layers from 2010 to 2100 in 5-year steps are first converted to an array of annual (interpolated) ENA. This annual ENA array is then used to drive relaxation. The first ten years should be identical to the last ten years of the relaxation over the historical period (as a sense-check), with the scenarios beginning to diverge only after 2020.

Each scenario's results are shown in three ways: a graph of the losses of ENA and endemic species from 2020-2100, expressed as percentages of the levels in 2020; a map of extinctions projected to take place between 2020 and 2100; and a summary of the relative numbers of endemics remaining in different selected years.

```{r relax_to_2100, warning=FALSE, cache=TRUE, echo=FALSE}
aim_files <- c("ENA-AIM-RCPref_SSP2_BID-12Feb2020_FinalMask.nc",
                 "ENA-AIM-RCPref_SSP2_CRJ-12Feb2020_FinalMask.nc",
                 "ENA-AIM-RCPref_SSP2_GDN-12Feb2020_FinalMask.nc",
                 "ENA-AIM-RCPref_SSP2_HPR-12Feb2020_FinalMask.nc",
                 "ENA-AIM-RCPref_SSP2_NRS-12Feb2020_FinalMask.nc",
                 "ENA-AIM-RCPref_SSP2_THC-12Feb2020_FinalMask.nc",
                 "ENA-AIM-RCPref_SSP2_WAC-12Feb2020_FinalMask.nc")
scenarios <- c("BID", "CRJ", "GDN", "HPR", "NRS", "THC", "WAC")
shortnames <- gsub("ENA-AIM-RCPref_SSP2_", "aslu_AIM", aim_files)
pre2010 <- historic_annual_area_aim[,,c(1:1160)] # 850 to 2009

for (i in 1:length(aim_files)){
  print(paste(Sys.time(), "Processing", aim_files[i], "..."))
  slu <- stack(file.path("output/ena", aim_files[i]))
  names(slu) <- seq(2010, 2100, by = 5)
  aslu <- interpolateTemporal(slu, xin=seq(2010, 2100, by=5), xout=seq(2010, 2100, by=1), 
                              outdir="figs", prefix=paste("AIM_", scenarios[i], sep=""), progress=TRUE, 
                                   writechange=FALSE, returnstack=TRUE, overwrite=TRUE)
  if (require(ncdf4)) {	
   rnc <- writeRaster(aslu, filename=file.path("output/aslu", paste("AIM_", shortnames[i], sep="")), format="CDF", overwrite=TRUE) 
  }

  # convert aslu to an array and save it
  aslu_array <- abind(pre2010, as.array(aslu))
  saveRDS(aslu_array, file.path("output/arrays/", paste(aim_files[i], ".Rds", sep="")))
  area_aslu  <- apply(aslu_array, MARGIN=3, sum, na.rm=TRUE)

  # relax using aslu array
  relaxation_scenario <- relax(natarea=aslu_array, years=c(850:2100), wanted=c(2010:2100), rsr_2015=rsr_array, k=this.k, z=this.z)
  S_t_scen <- relaxation_scenario$S_t
  saveRDS(S_t_scen, file.path("output/arrays/", paste(shortnames[i], ".Rds", sep="")))

  # Calculate numbers of endemic species and area, do percentage plots, and save time series
  endemics_scen <- apply(S_t_scen, MARGIN=3, sum, na.rm=TRUE)
  saveRDS(endemics_scen, file.path("output/endemics", paste(shortnames[i], ".Rds", sep="")))

  percentage_plot(A=aslu_array[,,c(1161:1251)], S=S_t_scen[,,c(1:91)], year=c(2010:2100), main_stem=shortnames[i])
  extinctions_map(S=S_t_scen, year=c(2010:2100), between=c(2010, 2100), main_stem=shortnames[i])
  scen_totals <- get_totals(S = endemics_scen, year = c(2010:2100))
  cat(paste(shortnames[i], "2020-2100: extinction summary\n"))
  print(scen_totals)
}
```



## Put results from scenarios into a data frame and plot them

The results are most naturally expressed as cumulative numbers of extinctions over time, on a relative scale.

```{r, harvest-results, echo=FALSE, warning=FALSE, cache=TRUE}
aim_projections <- data.frame(year = c(2010:2100))
aim_projections$BID <- readRDS(file.path("output/endemics","aslu_AIMBID-12Feb2020_FinalMask.nc.Rds"))
aim_projections$CRJ <- readRDS(file.path("output/endemics","aslu_AIMCRJ-12Feb2020_FinalMask.nc.Rds"))
aim_projections$GDN <- readRDS(file.path("output/endemics","aslu_AIMGDN-12Feb2020_FinalMask.nc.Rds"))
aim_projections$HPR <- readRDS(file.path("output/endemics","aslu_AIMHPR-12Feb2020_FinalMask.nc.Rds"))
aim_projections$NRS <- readRDS(file.path("output/endemics","aslu_AIMNRS-12Feb2020_FinalMask.nc.Rds"))
aim_projections$THC <- readRDS(file.path("output/endemics","aslu_AIMTHC-12Feb2020_FinalMask.nc.Rds"))
aim_projections$WAC <- readRDS(file.path("output/endemics","aslu_AIMWAC-12Feb2020_FinalMask.nc.Rds"))

aim_projections <- subset(aim_projections, year >= 2020)

for(i in 2:8){
  aim_projections[,i] <- aim_projections[1,i] - aim_projections[,i]
} 

toll <- max(aim_projections[,c(2:8)])
aim_projections[,c(2:8)] <- 100*aim_projections[, c(2:8)]/toll

plot(BID ~ year, data=aim_projections, col="red", type="l", ylim=c(0,100),
     ylab="Extinctions since 2020 (relative)", xlab="Year", cex.lab=1.3)
text(2020, 95, adj=0, "Key to scenarios:")
text(2020, 90, adj=0, "NRS", col="blue")
text(2020, 85, adj=0, "THC", col="darkblue")
text(2020, 80, adj=0, "CRJ", col="orange")
text(2020, 75, adj=0, "WAC", col="violet")
text(2020, 70, adj=0, "BID", col="red")
text(2020, 65, adj=0, "GDN", col="yellow")
text(2020, 60, adj=0, "HPR", col="green")

lines(CRJ ~ year, data=aim_projections, col="orange", type="l")
lines(GDN ~ year, data=aim_projections, col="yellow", type="l")
lines(HPR ~ year, data=aim_projections, col="green", type="l")
lines(NRS ~ year, data=aim_projections, col="blue", type="l")
lines(THC ~ year, data=aim_projections, col="darkblue", type="l")
lines(WAC ~ year, data=aim_projections, col="violet", type="l")


saveRDS(aim_projections, "output/endemics/aim_projections.Rds")
```

## Package version info

```{r, session-info, echo=FALSE, cache=FALSE}
utils::sessionInfo()
```
