---
title: "Using_AIM_data"
author: "Andy Purvis"
date: "28/02/2021"
output:
  html_document: 
    df_print: paged
    code_folding: hide
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
  pdf_document:
    df_print: kable
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Using the IAM data

The IMAGE data had the best fit to LUH2, but an odd pattern of land conversion in the years immediately following 2020 gave some highly non-intuitive results. Of the other three IAMs, AIM has the best fit to LUH2. This markdown therefore harmonizes the AIM scenario data to LUH2 and performs the dynamic species-area analysis.

## Reproducibility

The checkpoint package is used to aid reproducibility.

```{r, load, echo=FALSE, cache=FALSE}
library(checkpoint)
checkpoint("2021-02-16")

start.time <- Sys.time()
library(raster)
library(scales)
library(abind)
library(gplots)
library(colorspace)
library(maps)
library(ncdf4) # package for netcdf manipulation, suggested by https://rpubs.com/boyerag/297592
library(rgdal) # package for geospatial analysis, suggested by https://rpubs.com/boyerag/297592
library(ggplot2) # package for plotting, suggested by https://rpubs.com/boyerag/297592
source("00_functions.R")
source("convert_iam_data.R")
```

## Package version info

```{r, session-info, echo=FALSE, cache=FALSE}
utils::sessionInfo()
```
